---
title: R/lineup2 User Guide
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R/lineup2 User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
---

```{r knitr_options, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=4.5, dev="svg")
old_digits <- options("digits")$digits
options(digits=3)
```


[R/lineup2](https://github.com/kbroman/lineup2)
is an [R](https://www.r-project.org) package
providing tools for detecting and
correcting sample mix-ups between two sets of measurements, such as
between gene expression data on two tissues. It's a revised
version of [lineup](https://github.com/kbroman/lineup), to be more
general and not so closely tied to the [R/qtl](https://rqtl.org)
package.

We will illustrate some of the tools in this User Guide. We will use
the example data included in the package `lineup2ex`. These are a
subset of data from [Broman et al.
(2015)](https://doi.org/10.1534/g3.115.019778) (see also [Tian et al.
2015](https://doi.org/10.1534/genetics.115.179432)), and concern gene
expression microarray data on about 500 mice in two tissues,
gastrocnemius muscle (abbreviated "gastroc") and pancreatic islets
(abbreviated "islet"). While the original data included 40,000 gene
expression measurements, the example data consider just a subset of
200 genes: 100 chosen to be highly correlated between the two tissues,
and 100 others chosen at random.

When we load the [lineup2](https://github.com/kbroman/lineup2)
package, the data are immediately available, as `lineup2ex`, which is
a list containing the two tissues, each of which is a matrix, samples
&times; genes.

```{r load_package}
library(lineup2)
gastroc <- lineup2ex$gastroc
islet <- lineup2ex$islet
```

We seek to identify sample mix-ups: mislabeled rows in one or the
other data sets. One approach to do this is to first identify a
subset of correlated columns, and then to look at the correlations
between rows, for that subset of columns.

### Aligning rows or columns

To calculate the correlations between columns, we first need to align
the rows between the two matrices. We can do this with the function
`align_matrix_rows()`. There is a similar function
`align_matrix_cols()` for aligning columns.

```{r align_rows}
aligned <- align_matrix_rows(gastroc, islet)
```

The original matrices have `r nrow(gastroc)` and `r nrow(islet)` rows.
The output of `align_matrix_rows()` is a list with two matrices, where
the original matrices are reduced to their common rows, based on
their row names. The rows are also placed in the same order.
Here, each matrix in the output has `r nrow(aligned[[1]])`
rows, which is the number of samples in common between the two
tissues.

We could use `align_matrix_cols()` to align the columns (subsetting to
the common column names and putting them in the same order), but here they
are already aligned, and so no change occurs.

```{r align_cols}
aligned <- align_matrix_cols(aligned[[1]], aligned[[2]])
```

### Correlations between matrices

With the rows now aligned, we can now calculate the correlation in
gene expression between the two tissues, to identify genes that will
be useful for measuring sample similarity.

The function `corr_betw_matrices()` is used to calculate the
correlation between the columns in one matrix and the columns in
another matrix. There are several options; here we will use `what="paired"`
to get just the correlations between the paired columns.

```{r calc_paired_corr}
paired_corr <- corr_betw_matrices(aligned[[1]], aligned[[2]])
```

The result is a vector of `r ncol(aligned[[1]])` correlations, for the
pairs of columns. Here is a plot of the sorted correlations.

```{r to_reset_par, include=FALSE}
old_mar <- par("mar")
old_mfrow <- par("mfrow")
old_las <- par("las")
```

```{r plot_paired_corr}
par(mar=c(4.1, 5.1, 1.1, 1.1), las=1)
plot(sort(paired_corr), ylab="Correlations between column pairs, sorted", pch=16)
abline(h=0, lty=2, col="gray60")
```

Because of the way the genes were selected, there are 100 genes
with between-tissue correlation scattered around 0, and then another
100 genes with correlations > 0.75. We will grab that top 100 for
assessing sample similarity.

```{r select_top_genes}
selected_genes <- which(paired_corr > 0.75)
```

### Correlations/distances between samples

We can now calculate the similarity between the samples using these
selected genes. The simplest way to do so is by again using
`corr_betw_matrices()`, but with the transposed matrices so that the
samples become the columns.

```{r sample_correlations}
corr_betw_samples <- corr_betw_matrices(t(gastroc[,selected_genes]),
                                        t(islet[,selected_genes]), what="all")
```

The output is a `r nrow(gastroc)` &times; `r nrow(islet)` matrix, with
the (i,j)th element being the correlation between the ith row of
`gastroc` and the jth row of `islet`.

Note that there is a related function `dist_betw_matrices()`
to calculate the Euclidean or mean absolute distance between samples.
There is also `dist_betw_arrays()` for arrays with >2 dimensions.
Unlike `corr_betw_matrices()`, these functions look for distance
between rows not columns.


### Summarize correlations/distances

The first thing to do with the resulting similarity matrix is to plot
the values. The function `hist_self_nonself()` plots histograms of the
self-self correlations and the self-nonself correlations.

```{r hist_self_nonself, fig.height=6, dev="png"}
hist_self_nonself(corr_betw_samples, xlabel="correlation")
```

The self-self correlations are mostly > 0.7, while the
self-nonself correlations are mostly < 0.7. The bimodal distribution
of the self-nonself correlations is due to sex differences. Same-sex
mice generally show positive correlations, while opposite-sex mice
generally show negative correlations.




### Identify sample mix-ups

There are a few self-self correlations that are small and even
negative. These are the sample mix-ups we are looking for. To identify
the problem samples, we want to pull out the individual values,
and identify the maximum values in each row and column.

You can pull out the self-self correlations with the function `get_self()`.
These are the cases where the row and column names match.

```{r self_values}
self <- get_self(corr_betw_samples)
```

There is also a function `get_nonself()` for pulling out the non-self values.

Note that there are `r sum(is.na(self))` missing values, where a
sample is present in only one of the two tissues:

```{r missing_self}
self[is.na(self)]
```

Here are the smallest eight values. As we will see, the first five of
these are the real problems.

```{r self_sorted}
sort(self)[1:8]
```


The samples with low self-self correlation are likely mix-ups. To
identify the correct labels, we can look for the largest correlation
in each row and column of the similarity matrix. To do this, we use
the function `get_best()`. We use the argument `get_min=FALSE` since
we are looking for the maximum here, and we use the argument
`dimension` to control whether to look within each row (gastroc) or
within each column (islet).

```{r best_by_row_and_col}
best_byrow <- get_best(corr_betw_samples, get_min=FALSE, dimension="row")
best_bycol <- get_best(corr_betw_samples, get_min=FALSE, dimension="column")
```

We can plot the self-self correlations against the maximum
correlations by row and column. (There is a bit of effort to get
labels on the points.)

```{r plot_self_v_best}
par(mfrow=c(1,2), mar=c(4.1, 4.1, 1.6, 0.6), las=1)
plot(self, best_byrow, xlab="Self-self correlation", ylab="Best islet correlation",
     main="Gastroc samples", pch=16, xlim=c(-0.3, 1.0), ylim=c(0.6, 1.0))
label <- best_byrow-self > 0.2
text(self[label]+0.03, best_byrow[label], names(self)[label],
     adj=c(0,0.5), cex=0.8)

plot(self, best_bycol, xlab="Self-self correlation", ylab="Best gastroc correlation",
     main="Islet samples", pch=16, xlim=c(-0.3, 1.0), ylim=c(0.6, 1.0))
label <- best_bycol-self > 0.3
text(self[label]+0.03, best_bycol[label], names(self)[label],
     adj=c(0,0.5), cex=0.8)
label <- best_bycol-self > 0.2 & best_bycol-self < 0.3
text(self[label]-0.03, best_bycol[label], names(self)[label],
     adj=c(1,0.5), cex=0.8)
```

Most samples have a large self-self correlation that matches the
maximum in the row and column. There are four gastroc samples that
have low self-self correlation but where there is another large value
in that row, and there are five problem islet samples.

The function `get_problems()` can be used to get a summary of
potential problems. Looking by row or by column, you can pull out the
samples where the best distance exceeds the self-self distance by more
than some threshold.

Looking by row, and using a threshold of 0.2, we see these samples:

```{r problems_byrow}
get_problems(corr_betw_samples, threshold=0.2, get_min=FALSE)
```

This shows the four problems in the left panel of the above figure,
and that they constitute two apparent sample swaps (Mouse3655 &harr;
Mouse3659 and Mouse3598 &harr; Mouse3599).

Looking by column, again using a threshold of 0.2, we see the same
results plus one more:

```{r problems_bycol}
get_problems(corr_betw_samples, threshold=0.2, get_min=FALSE, dimension="column")
```

We again see the Mouse3655 &harr;
Mouse3659 and Mouse3598 &harr; Mouse3599 pairs, and also see that
Mouse3296 sample in islet looks to correspond to Mouse3295 (gastroc).

Note that the function `which_best()` can be used to provide the
sample IDs that correspond to the results of `get_best()`; sort of
like `which.max()` vs `max()`. This is another way to see that islet
Mouse3296 looks like Mouse3295.

```{r which_best_islet3296}
which_best(corr_betw_samples, get_min=FALSE, dimension="column")["Mouse3296"]
```

Another way to look at these problems is to plot the values in each
column. We will point to the maximum in red and the self value in green.

```{r plot_six_samples, fig.height=6.5}
samples <- sort(c(names(self)[best_bycol-self > 0.2], "Mouse3295"))
par(mfrow=c(3,2), las=1, mar=c(4.1, 4.1, 2.1, 0.6))
rn <- rownames(corr_betw_samples)
green <- "#2ecc40"
red <- "#ff4136"
for(sample in samples) {
    plot(corr_betw_samples[,sample], xlab="Index", ylab="Correlation",
         main=sample, pch=16, ylim=range(corr_betw_samples))
    mx <- which.max(corr_betw_samples[,sample])
    points(mx, corr_betw_samples[mx,sample], pch=16, col=red)
    text(mx-8, corr_betw_samples[mx, sample], names(mx), adj=c(1, 0.5))
    points(which(rn==sample), corr_betw_samples[sample,sample], pch=16, col=green)
}
```

From these results, we can see that Mouse3598 and Mouse3599 look to be
switched in one or the other tissue, as are Mouse3655 and Mouse3659.
From these data on their own, we can't tell which tissue is the
problem, but the original data include gene expression microarrays for
four other tissues, and by comparisons to those tissues, it turns out
that Mouse3655/Mouse3659 were mixed up in gastroc, while
Mouse3598/Mouse3599 were mixed up in islet. See [Broman et al.
(2015)](https://doi.org/10.1534/g3.115.019778), [Fig.
4](https://www.g3journal.org/content/ggg/5/10/2177/F4.large.jpg).

Samples Mouse3295 and Mouse3296 look correct in gastroc, and Mouse3295
looks correct in islet, but Mouse3296 islet looks like Mouse3295
gastroc. The sample labelled Mouse3296 in islet is really an
unintended biological replicate of sample Mouse3295.


### Second-best samples

As further support to show that the best-correlated samples correspond
to the correct labels, it can be useful to consider the second-best
sample in each row and column. If the best-correlated sample is not
much more correlated than the second-best, than the support is not so
strong, for the best sample to be the true one.

To pull out the second-best correlated sample, use the function
`get_2ndbest()`, which has the same arguments as `get_best()`.

```{r second_best_by_row_and_col}
secbest_byrow <- get_2ndbest(corr_betw_samples, get_min=FALSE, dimension="row")
secbest_bycol <- get_2ndbest(corr_betw_samples, get_min=FALSE, dimension="column")
```

We then can plot the second-best samples versus the best samples, and
we will point to the problem samples. The four problem samples in
gastroc, and the five problem samples in islet, are highlighted in red.

```{r sec_best_vs_best}
par(mfrow=c(1,2), mar=c(4.1, 4.1, 1.6, 0.6), las=1)
plot(best_byrow, secbest_byrow, xlab="Best islet correlation",
     ylab="Second best islet correlation",
     main="Gastroc samples", pch=16, xlim=c(0.5, 1), ylim=c(0.5, 1))
label <- best_byrow-self > 0.2
points(best_byrow[label], secbest_byrow[label], pch=16, col=red)
abline(0,1, lty=2)

plot(best_bycol, secbest_bycol, xlab="Best gastroc correlation",
     ylab="Second best gastroc correlation",
     main="Islet samples", pch=16, xlim=c(0.5, 1), ylim=c(0.5, 1))
label <- best_bycol-self > 0.2
points(best_bycol[label], secbest_bycol[label], pch=16, col=red)
abline(0,1, lty=2)
```

Values away from the diagonal have good support for the "best" sample
being the correct one. This includes all of the problem samples (in red).

Note that there is `which_2ndbest()` function, similar to
`which_best()` but pulling out the second-closest sample.


### References

Broman KW, Keller MP, Broman AT, Kendziorski C, Yandell BS, Sen Ś,
Attie AD (2015) Identification and correction of sample mix-ups in
expression genetic data: A case study. G3 5:2177-2186
[doi:10.1534/g3.115.019778](https://doi.org/10.1534/g3.115.019778)

Tian J, Keller MP, Oler AT, Rabaglia ME, Schueler KL, Stapleton DS,
Broman AT, Zhao W, Kendziorski C, Yandell BS, Hagenbuch B, Broman KW,
Attie AD (2015) Identification of the bile acid transporter Slco1a6 as
a candidate gene that broadly affects gene expression in mouse
pancreatic islets. Genetics 201:1253-1262
[doi:10.1534/genetics.115.179432](https://doi.org/10.1534/genetics.115.179432)


```{r reset_par_and_options, include=FALSE}
par(mar=old_mar, mfrow=old_mfrow, las=old_las)
options(digits=old_digits)
```
