---
title: <center> Package JumpTest usage tips
author: |
  <center> Kaiqiao Li, and Pei Fen Kuan
  <center> Department of Applied Mathematics and Statistics, Stony Brook University
  <center> Stony Brook, NY 11790
date: <center>`r Sys.Date()`
output: rmarkdown::html_vignette
bibliography: JumpTest.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Overview

This vignette provides an introduction to nonparametric tests to financial jump detection. R package `JumpTest` provided both simulation methods and test methods with multiple tests correction. The package can be loaded with command:

```{r}
library(JumpTest)
```


## Workflow: Meta Analysis

This package provided analysis on high frequency stock markets. You can simply simulate the data using the simulation codes we provided.
For example you can simulate a stoachasitc volatility with one factor and jump (SV1FJ) model [@dumitru2012identifying] for 1200 days using the following code.

```{r}
M <- 390*60
m <- 1200
pric <- SV1FJ(M,m,3,.2,.03,.5,0,.125,-.1,-.62)
lp <- c(3,pric[,1])
lr <- diff(lp[seq(1,M*m+1,60)])
logreturn <- matrix(lr,ncol=m)
```

Here we simulated our data by seconds, while we grabbed the minute level data and saved the log returns per day by columns. Then we can used the following code to obtain the p-values for three methods (BNS [@barndorff2006limit], Amed, and Amin [@andersen2012jump]).

```{r}
pmat <- pcombine(logreturn,c('BNS','Amed','Amin'))
```

Then we can obtain the p-value matrix for the three methods. We can do further p-values pooling and FDR control via BH [@benjamini1995controlling, @yen2013testing] procedure. We provided six pooling methods, four for independent [@won2009choosing, @chang2013meta] and two for dependent [@chang2013meta] strucure. We suggested Stouffer's dependent (SD) structure.

```{r}
res <- ppool(pmat)
pvalue <- res@pvalue
pvalue_adjusted <- res@adjp
```

## Conclusion and Ongoing Work

R package `JumpTest` provides effective tools to detect jump in high frequency stock market data. It can efficiently identify intervals that contain jumps. In the future we may develop methods that can locate the exact location of jumps.

## REFERENCES