---
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sobol sequences with better two-dimensional projections}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
---

Sobol Sequence
--------------

R implementation of
S. Joe and F. Y. Kuo,
"Constructing Sobol sequences with better two-dimensional projections",
SIAM J. Sci. Comput. 30, 2635-2654 (2008).
The implementation is based on the data file new-joe-kuo-6.21201
<http://web.maths.unsw.edu.au/~fkuo/sobol/>.
        

Get avairable dimension number of Sobol Sequence.
```{r}
library(SobolSequence)
sobolSequence.dimMinMax()
```

Get avairable F2 dimension number of Sobol Sequence.
```{r}
sobolSequence.dimF2MinMax(10)
```

And get points. Each row of returned matrix contains an s-dimensional point.
```{r}
s <- 4
m <- 10
c <- 2^m
mat <- sobolSequence.points(dimR=s, dimF2=m, count=c)
mat[1,]
```

Get digital shifted points. 
```{r}
s <- 4
m <- 10
c <- 2^m
mat <- sobolSequence.points(dimR=s, dimF2=m, count=c, digitalShift=TRUE)
mat[1,]
```
