---
title: "Computing your Eddington number using the `eddington` package."
author:
  "Paul W. Egeler, M.S."
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eddington}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

# Putting this here too so that we don't have pkg startup messages.
library(dplyr)
```

## Introduction

One statistic that cyclists might be interested to track is their Eddington
number. The Eddington number for cycling, _E_, is the maximum number where a
cyclist has ridden _E_ miles in _E_ days. So to get a number of 30, you need to
have ridden 30 miles or more on 30 separate days.

This package allows the user to compute Eddington numbers and more. For example,
users can determine if a specific Eddington number is satisfied or how many
rides of the appropriate distance are needed to increment their Eddington
number. It also contains **simulated** data to demonstrate the package use.

## The Data

Loading the simulated data is simple. Let's take a quick look at the first few lines.

```{r setup}
library(eddington)
head(rides)
```

First, we need to establish the granularity of the data. As you can see above,
there are at least two entries for 2009-01-18. Since this data simulates a rider
who tracked each individual ride, there could be more than one ride per day in
this dataset. Therefore, we need to transform the data to aggregate on day.

```{r xform}
library(dplyr)

days <- rides %>%
  group_by(ride_date) %>%
  summarize(n = n(), total = sum(ride_length))

head(days)
```

Let's just take a quick peek at the summary stats:

```{r summary}
summary(days)
```


### Histogram of Daily Mileages

This plot provides a histogram of daily mileages. Note the summary Eddington
number is in dark red---we'll see how that's calculated in the next section.

```{r, echo=FALSE}
hist(
  as.integer(days$total), 
  breaks = 30, 
  main = "Histogram of Daily Mileages", 
  xlab = "Miles"
)

abline(v = E_num(days$total), col = "darkred")

legend(
  "topright",
  legend = "Eddington Number",
  col = "darkred",
  bty = "n",
  lty = 1L
)
```

## Computing Eddington Numbers

To compute the Eddington number, we use the `E_num()` function like so:

```{r enum}
E_num(days$total)
```

### Cumulative E

To see how the Eddington number progressed over the year, use `E_cum()`. It can
be useful to add the vector as a new column onto the existing dataset:

```{r ecum}
days$E <- E_cum(days$total)

head(days)
```

It might be more interesting to see that graphically:

```{r needle, echo=FALSE}
E <- E_num(days$total)

E_contribs <- days[days$total >= E,]


plot(
  y = days$total,
  x = days$ride_date,
  type = "h",
  main = "Ride Mileages in 2009",
  xlab = "Ride Day",
  ylab = "Total Miles",
  bty = "n",
  ylim = c(0, 90)
)


lines(
  y = c(0, days$E),
  x = c(as.Date("2009-01-01"), days$ride_date),
  type = "s",
  col = "darkred"
)

abline(h = E, lty = 2L, col = "darkred")

text(
  E_contribs[,c("ride_date","total")],
  labels = as.integer(E_contribs[["total"]]),
  pos = 3,
  cex = 0.7
  )

legend(
  "topleft",
  title = "Eddington Number",
  legend = c("Cumulative", "Summary"),
  col = "darkred",
  bty = "n",
  lty = c(1L, 2L)
)
```

## Addtional Functionality

### Incrementing to the Next Eddington Number

So now that we know that the summary Eddington number was 29 for the year, let's
see how many more rides of length 30 or greater that we would have needed to
increment the _E_ to 30.

```{r enext}
E_next(days$total)
```

### Stretch Goals

An ambitious rider might be interested to see the number of rides required to
reach a stretch goal. Say, how many more rides would have been needed to reach
an _E_ of 50? For that, we use `E_req()`, which stands for "required."

```{r ereq}
E_req(days$total, 50)
```

### Check if a Dataset Satisfies an Arbitrary _E_

We could also check to see if we've gotten to 30 by using `E_sat()`, which
stands for "satisfies."

```{r esat}
E_sat(days$total, 30)
```

## Conclusion

The text above should give you a good start in using the `eddington` package.
Although this package was developed with bicycling in mind, it has applications
for other users as well. The Eddington number is a specific application of
computing the side length of a Durfee square. Another application is the Hirsch
index, or _h_-index, which a popular number in bibliometrics.
