---
title: "Rdca Package"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
csl: apa-6th-edition.csl
vignette: >
  %\VignetteIndexEntry{Rdca Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## `Rdca`

`Rdca` is a Decline Curve Analysis (DCA) package for oil and gas reservoirs. It generates a table of rate, cumulative, nominal decline rate, and derivative of loss-ratio over time in a data frame format. It also provides an optimization tool to fit a DCA model on production data. The package currently supports Arps 'exponential', 'harmonic', 'hyperbolic', and 'modified_hyperbolic' models [@Arps1945; @Robertson1988].

DCA predictions are generated using three different functions: `decline_param()`, `decline_time()`, and `decline_predict()`. There are two more functions for the curve fitting exercises: `decline_fit_param()`, and `decline_fit()`.

## `decline_param() arguments`

+ `input_unit`: A unit system for input parameters, either 'Field' or 'SI'.
+ `output_unit`: A unit system for output predictions, either 'Field' or 'SI'.
+ `fluid`: Type of fluid, either 'gas' or 'oil'.
+ `model`: Arps DCA model.
+ `qi`: Arps model parameter, initial rate.
+ `Di`: Arps model parameter, nominal decline rate.
+ `b`: Arps model parameter, decline constant.
+ `Dt`: Arps modified_hyperbolic model parameter, terminal nominal decline rate.
+ `q_abnd`: Abandonment rate, an optional input parameter. If assigned a value, the model predicts time to reach to the abandonment rate, and also the estimated ultimate recovery (EUR).

## `decline_time() arguments`

+ `x`: A vector or sequence of times/dates.
+ `unit`: A unit system for input vector x.

## `decline_predict() arguments`

+ `decline_lst`: A list of decline parameters of class 'decline'.
+ `time_lst`: A list of decline time of class 'time/date'.


## `decline_fit_param() arguments`

+ `input_unit`: A unit system for input parameters, either 'Field' or 'SI'.
+ `output_unit`: A unit system for output predictions, either 'Field' or 'SI'.
+ `fluid`: Type of fluid, either 'gas' or 'oil'.
+ `model`: Arps DCA model.
+ `fit_data`: Fitting data, either 'rate' or 'cum'.
+ `prod_data`: A numeric vector of fitting data.
+ `initial_param`: A vector of initial estimates for the Arps DCA model parameters.
+ `lower`: A vector of lower bounds for the Arps DCA model parameters, optional. See package `minpack.lm` for more information. 
+ `upper`: A vector of upper bounds for the Arps DCA model parameters, optional. See package `minpack.lm` for more information. 
+ `control`: An optional list of control settings in the optimization toolbox. See package `minpack.lm` for more information. 

## `decline_fit() arguments`

+ `decline_fit_lst`: A list of decline fit parameters of class 'decline_fit'.
+ `time_lst`: A list of decline time of class 'time/date'.


## `Units` for input parameters

The input_unit is either 'SI' or 'Field'. Depending on the input_unit system, the following units are used for the input parameters:

+ `qi`: Arp's decline parameter, a numeric value. Depending on the 'input_unit', 'fluid' type, and also the decline_time() 'unit' parameter, it has different units. 'm3/day' for gas production in 'SI' unit with daily data, 'm3/month' for gas production in 'SI' unit with monthly data, 'MSCF/day' for gas production in 'Field' unit with daily data, 'MSCF/month' for gas production in 'Field' unit with monthly data, 'm3/day' for oil production in 'SI' unit with daily data, 'm3/month' for oil production in 'SI' unit with monthly data, 'bbl/day' for oil production in 'Field' unit with daily data, and 'bbl/month' for oil production in 'Field' unit with monthly data.
+ `Di`: Arp's nominal decline parameter, a numeric value in '1/day', '1/month', or '1/year' depending on the decline_time() 'unit' parameter.
+ `b`: Arp's decline parameter. It is zero for the 'exponential' model, one for the 'harmonic' model, and a value between zero and one for the 'hyperbolic' model. For unconventional reservoirs, b values more than one are also reported.
+ `Dt`: Arp's 'modified_hyperbolic' nominal terminal decline parameter, a numeric value in '1/day', '1/month', or '1/year' depending on the decline_time() 'unit' input.
+ `q_abnd`: Abandonment rate, a numeric value defaulted to NULL. If present, the model predicts the time to reach to the abandonment rate and also the estimated ultimate recovery (EUR) till the abandonment time. It has the same unit as 'qi'.
+ `unit`: time/date unit for the x vector. For a vector of time values, the unit is 'day', 'month', or 'year'. For a sequence of dates, the unit is 'date'. Daily parameters must be used with the 'date' unit in the DCA.

## Outputs

+ `Date`: Date
+ `Time`: Time
+ `q`: Production rate
+ `Q`: Cumulative production
+ `D`: Nominal decline rate (1/loss-ratio)
+ `Beta`: Loss-ratio derivative wrt time
+ `time_abnd`: Time to reach to the abandonment rate (if 'abnd_rate' is provided)
+ `EUR`: Estimated ultimate recovery at abandonment rate (if 'abnd_rate' is provided)


## A note on `Di`

Arps `Di` parameter is the nominal decline rate in '1/day', '1/month', or '1/year'. It represents the rate of decline at a specific time. However, the rate of decline is often expressed for a particular period (usually one year) and is called the effective decline rate `(di)` in '1/year'. The relationship between `Di` and `di` for Arps models are as follows:

- Exponential 

$$D_{i} = \frac{-log(1 - d_{i})}{365}\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\: \frac{1}{day}$$<br><br>


$$D_{i} = \frac{-log(1 - d_{i})}{12}\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\: \frac{1}{month}$$<br><br>


$$D_{i} = -log(1 - d_{i}) \:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\: \frac{1}{year}$$<br><br>

- Harmonic 

$$D_{i} = \frac{d_{i}}{(1 - d_{i}) . 365}\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\: \frac{1}{day}$$<br><br>


$$D_{i} = \frac{d_{i}}{(1 - d_{i}) . 12}\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\: \frac{1}{month}$$<br><br>


$$D_{i} = \frac{d_{i}}{(1 - d_{i})}\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\: \frac{1}{year}$$<br><br>


- Hyperbolic 

$$D_{i} = \frac{1}{b}.\frac{[(1 - d_{i})^{-b} - 1]}{365}\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\: \frac{1}{day}$$<br><br>


$$D_{i} = \frac{1}{b}.\frac{[(1 - d_{i})^{-b} - 1]}{12}\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\: \frac{1}{month}$$<br><br>


$$D_{i} = \frac{1}{b}.[(1 - d_{i})^{-b} - 1]\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\: \frac{1}{year}$$<br><br>


## `Installation`

The `Rdca` can be installed from [CRAN](https://CRAN.R-project.org) with:


```{r, eval = FALSE, warning = FALSE}
install.packages("Rdca")
```

## `Arps Exponential Examples`

#### Example 1:

```{r, fig.width = 6, fig.height= 8, fig.align = "center", warning = TRUE}
library(Rdca)
library(magrittr)
library(ggplot2)
library(ggpubr)

dcl_param_exp <- decline_param(input_unit = "Field", output_unit = "Field", fluid = "oil", 
                               model = "exponential", qi = 1000, Di = 0.0015, b = 0, q_abnd = NULL)

dcl_param_exp

decline_time_exp <- decline_time(c(1:7300), unit = "day")   

str(decline_time_exp)

decline_predict_exp <- decline_predict(dcl_param_exp, decline_time_exp)

head(decline_predict_exp, 10)

p1 <- decline_predict_exp %>% ggplot(aes(x = `Time_(day)`, y = `q_(bbl/day)`)) + 
  geom_point(color = "green4") +  
  theme_bw()

p2 <- decline_predict_exp %>% ggplot(aes(x = `Time_(day)`, y = `Q_(bbl)`)) + 
  geom_point(color = "green4") +
  theme_bw()

exp_plots <- ggarrange(p1, p2, ncol = 1, nrow = 2, align = "v")

exp_plots

```

#### Example 2:

```{r, fig.width = 6, fig.height= 8, fig.align = "center", warning = TRUE}
library(Rdca)
library(ggplot2)
library(ggpubr)

dcl_param_exp <- decline_param(input_unit = "Field", output_unit = "SI", fluid = "gas", 
                               model = "exponential", qi = 75000, Di = 0.03, b = 0, q_abnd = 1000)

dcl_param_exp

decline_time_exp <- decline_time(c(1:360), unit = "month")   

str(decline_time_exp)

decline_predict_exp <- decline_predict(dcl_param_exp, decline_time_exp)

head(decline_predict_exp, 10)

time_abnd <- decline_predict_exp$`time_abnd_(months)`[1]
EUR <- decline_predict_exp$`EUR_(m3)`[1]

p1 <- decline_predict_exp %>% ggplot(aes(x = `Time_(month)`, y = `q_(m3/month)`)) +
  geom_point(color = "red") +
  geom_vline(aes(xintercept = `time_abnd_(months)`), linetype = 2) + 
  annotate(geom = "text", x = time_abnd + 10, y = 5e5, label = "time_abnd", angle = 90, 
           color = "blue") +
  geom_hline(aes(yintercept = 1e6 / 35.3147), linetype = 2) + 
  annotate(geom = "text", x = 50, y = 3e6 / 35.3147, label = "rate_abnd", color = "blue") +
  theme_bw()

p2 <- decline_predict_exp %>% ggplot(aes(x = `Time_(month)`, y = `Q_(m3)`)) +
  geom_point(color = "red") +
  geom_vline(aes(xintercept = `time_abnd_(months)`), linetype = 2) + 
  annotate(geom = "text", x = time_abnd + 10, y = 6e7, label = "time_abnd", angle = 90, 
           color = "blue") +
  geom_hline(aes(yintercept = `EUR_(m3)`), linetype = 2) + 
  annotate(geom = "text", x = 50, y = 1.02 * EUR, label = "EUR", color = "blue") +
  theme_bw()

exp_plots <- ggarrange(p1, p2, ncol = 1, nrow = 2, align = "v")

exp_plots

```

#### Example 3:

```{r, fig.width = 6, fig.height= 8, fig.align = "center", warning = TRUE}
library(Rdca)
library(ggplot2)

dcl_time_exp <- decline_time(1:1000, unit = "day")

set.seed(123)

prod_data <- 3000 * exp(-0.00234 * dcl_time_exp$t) + 50 * rnorm(1000)

field_data <- data.frame(time = dcl_time_exp$t, q = prod_data)

dcl_fit_param_exp <- decline_fit_param(input_unit = "Field", output_unit = "Field", 
                                       fluid = "oil", model = "exponential", 
                                       fit_data = "rate", prod_data = prod_data, 
                                       initial_param = c(3000, 0.001, 0), lower = NULL, 
                                       upper = NULL)

tibble::glimpse(dcl_fit_param_exp)

dcl_fit_exp <- decline_fit(dcl_fit_param_exp, dcl_time_exp)

tibble::glimpse(dcl_fit_exp)

names(attr(dcl_fit_exp, which = "nls.out"))

attr(dcl_fit_exp, which = "nls.out")$par

attr(dcl_fit_exp, which = "nls.out")$info

attr(dcl_fit_exp, which = "nls.out")$niter

attr(dcl_fit_exp, which = "nls.out")$deviance

dcl_predict_exp <- decline_predict(dcl_fit_exp, dcl_time_exp)

field_data %>% ggplot(aes(x = time, y = q)) +
  geom_point(color = "blue", shape = 21, size = 3) +
  geom_line(aes(x = `Time_(day)`, y = `q_(bbl/day)`), data = dcl_predict_exp,
            color = "red", size = 1) +
  labs(x = "Time (days)", y = "Rate (bbl/day)") + 
  theme_bw()

```



## `Arps Harmonic Examples`

#### Example 1:

```{r, fig.width = 6, fig.height= 8, fig.align = "center", warning = TRUE}
library(Rdca)
library(ggplot2)
library(ggpubr)

dcl_param_harm <- decline_param(input_unit = "SI", output_unit = "SI", fluid = "oil", 
                               model = "harmonic", qi = 1000, Di = 0.075, b = 1, q_abnd = 50)

dcl_param_harm

decline_time_harm <- decline_time(c(1:360), unit = "month")   

str(decline_time_harm)

decline_predict_harm <- decline_predict(dcl_param_harm, decline_time_harm)

head(decline_predict_harm, 10)

time_abnd <- decline_predict_harm$`time_abnd_(months)`[1]

EUR <- decline_predict_harm$`EUR_(m3)`[1]

p1 <- decline_predict_harm %>% ggplot(aes(x = `Time_(month)`, y = `q_(m3/month)`)) + 
   geom_point(color = "green4") +
  geom_vline(aes(xintercept = `time_abnd_(months)`), linetype = 2) + 
  annotate(geom = "text", x = time_abnd + 10, y = 250, label = "time_abnd", angle = 90, 
           color = "blue") +
  geom_hline(aes(yintercept = 50), linetype = 2) + 
  annotate(geom = "text", x = 50, y = 80, label = "rate_abnd", color = "blue") +
  theme_bw()

p2 <- decline_predict_harm %>% ggplot(aes(x = `Time_(month)`, y = `Q_(m3)`)) + 
  geom_point(color = "green4") +
  geom_vline(aes(xintercept = `time_abnd_(months)`), linetype = 2) + 
  annotate(geom = "text", x = time_abnd + 10, y = 35000, label = "time_abnd", angle = 90, 
           color = "blue") +
  geom_hline(aes(yintercept = `EUR_(m3)`), linetype = 2) + 
  annotate(geom = "text", x = 50, y = 1.04 * EUR, label = "EUR", color = "blue") +
  theme_bw()

harm_plots <- ggarrange(p1, p2, ncol = 1, nrow = 2, align = "v")

harm_plots

```


#### Example 2:

```{r, fig.width = 6, fig.height= 8, fig.align = "center", warning = TRUE}
library(Rdca)
library(ggplot2)

dcl_time_harm <- decline_time(1:360, unit = "month")

set.seed(1234)

prod_data <- 30000 / (1 + 0.02 * dcl_time_harm$t) + 500 * rnorm(360)   # rate

field_data <- data.frame(time = dcl_time_harm$t, q = prod_data)

field_data$Q <- cumsum(field_data$q)                              # cumulative

dcl_fit_param_harm <- decline_fit_param(input_unit = "SI", output_unit = "SI", fluid = "gas", 
                              model = "harmonic", fit_data = "cum", prod_data = field_data$Q, 
                              initial_param = c(40000, 0.01, 1), lower = NULL, upper = NULL,
                              control = list(maxiter = 100))

tibble::glimpse(dcl_fit_param_harm)

dcl_fit_harm <- decline_fit(dcl_fit_param_harm, dcl_time_harm)

tibble::glimpse(dcl_fit_harm)

names(attr(dcl_fit_harm, which = "nls.out"))

attr(dcl_fit_harm, which = "nls.out")$par

attr(dcl_fit_harm, which = "nls.out")$info

attr(dcl_fit_harm, which = "nls.out")$niter

attr(dcl_fit_harm, which = "nls.out")$deviance

dcl_predict_harm <- decline_predict(dcl_fit_harm, dcl_time_harm)

p_cum <- field_data %>% ggplot(aes(x = time, y = Q)) +
  geom_point(color = "blue", shape = 21, size = 3) +
  geom_line(aes(x = `Time_(month)`, y = `Q_(m3)`), data = dcl_predict_harm, 
            color = "red", size = 1) +
  labs(x = "Time (months)", y = "Cumulative Production (m3)") + 
  theme_bw()

p_cum

p_rate <- field_data %>% ggplot(aes(x = time, y = q)) +
  geom_point(color = "blue", shape = 21, size = 3) +
  geom_line(aes(x = `Time_(month)`, y = `q_(m3/month)`), data = dcl_predict_harm, 
            color = "red", size = 1) +
  labs(x = "Time (months)", y = "Production Rate (m3/month)") + 
  theme_bw()

p_rate

```


## `Arps Hyperbolic Examples`

#### Example 1:

```{r, fig.width = 6, fig.height= 8, fig.align = "center", warning = TRUE}
library(Rdca)
library(ggplot2)
library(ggpubr)

dcl_param_hyp <- decline_param(input_unit = "Field", output_unit = "Field", fluid = "gas", 
                               model = "hyperbolic", qi = 100000, Di = 0.0055, b = 0.85, 
                               q_abnd = 2000)

dcl_param_hyp

decline_time_hyp <- decline_time(seq(as.Date("2000/1/1"), as.Date("2030/12/31"), "days"), 
                                 unit = "date")   

str(decline_time_hyp)

decline_predict_hyp <- decline_predict(dcl_param_hyp, decline_time_hyp)

head(decline_predict_hyp, 10)

time_abnd <- decline_predict_hyp$`time_abnd_(days)`[1]

EUR <- decline_predict_hyp$`EUR_(MMSCF)`[1]

p1 <- decline_predict_hyp %>% ggplot(aes(x = `Time_(day)`, y = `q_(MSCF/day)`)) + 
   geom_point(color = "red") +
  geom_vline(aes(xintercept = `time_abnd_(days)`), linetype = 2) + 
  annotate(geom = "text", x = time_abnd + 200, y = 25000, label = "time_abnd", angle = 90, 
           color = "blue") +
  geom_hline(aes(yintercept = 2000), linetype = 2) + 
  annotate(geom = "text", x = 7500, y = 5000, label = "rate_abnd", color = "blue") +
  theme_bw()

p2 <- decline_predict_hyp %>% ggplot(aes(x = `Time_(day)`, y = `Q_(MMSCF)`)) + 
  geom_point(color = "red") +
  geom_vline(aes(xintercept = `time_abnd_(days)`), linetype = 2) + 
  annotate(geom = "text", x = time_abnd + 200, y = 4e4, label = "time_abnd", angle = 90, 
           color = "blue") +
  geom_hline(aes(yintercept = `EUR_(MMSCF)`), linetype = 2) + 
  annotate(geom = "text", x = 8000, y = 1.03 * EUR, label = "EUR", color = "blue") +
  theme_bw()

hyp_plots <- ggarrange(p1, p2, ncol = 1, nrow = 2, align = "v")

hyp_plots

```


#### Example 2:

```{r, fig.width = 6, fig.height= 8, fig.align = "center", warning = TRUE}
library(Rdca)
library(ggplot2)

dcl_time_hyp <- decline_time(1:10000, unit = "day")

set.seed(321)

prod_data <- 4500 / (1 + 0.002 * 0.834 * dcl_time_hyp$t) ^ (1 / 0.834) + 
  25 * rnorm(10000)   # rate

field_data <- data.frame(time = dcl_time_hyp$t, q = prod_data)

dcl_fit_param_hyp <- decline_fit_param(input_unit = "Field", output_unit = "Field", fluid = "gas", 
                              model = "hyperbolic", fit_data = "rate", prod_data = prod_data, 
                              initial_param = c(1000, 0.01, 1.0), lower = c(0, 1e-6, 1e-6),
                              upper = NULL, control = list(maxiter = 100))

tibble::glimpse(dcl_fit_param_hyp)

dcl_fit_hyp <- decline_fit(dcl_fit_param_hyp, dcl_time_hyp)

tibble::glimpse(dcl_fit_hyp)

names(attr(dcl_fit_hyp, which = "nls.out"))

attr(dcl_fit_hyp, which = "nls.out")$par

attr(dcl_fit_hyp, which = "nls.out")$info

attr(dcl_fit_hyp, which = "nls.out")$niter

attr(dcl_fit_hyp, which = "nls.out")$deviance

dcl_predict_hyp <- decline_predict(dcl_fit_hyp, dcl_time_hyp)

field_data %>% ggplot(aes(x = time, y = q)) +
  geom_point(color = "blue", shape = 21, size = 3) +
  geom_line(aes(x = `Time_(day)`, y = `q_(MSCF/day)`), data = dcl_predict_hyp,
            color = "red", size = 1) +
  labs(x = "Time (days)", y = "Rate (MSCF/day)") + 
  theme_bw()

```

## `Arps Modified_Hyperbolic Examples`

#### Example 1:

```{r, fig.width = 6, fig.height= 8, fig.align = "center", warning = TRUE}
library(Rdca)
library(ggplot2)
library(ggpubr)

dcl_param_mod_hyp <- decline_param(input_unit = "SI", output_unit = "Field", fluid = "oil", 
                               model = "modified_hyperbolic", qi = 1000, Di = 0.0055, b = 0.85, 
                               Dt = 0.0005, q_abnd = 5)
dcl_param_mod_hyp

decline_time_mod_hyp <- decline_time(seq(as.Date("2000/1/1"), as.Date("2030/12/31"), "days"), 
                                 unit = "date")   

str(decline_time_mod_hyp)

decline_predict_mod_hyp <- decline_predict(dcl_param_mod_hyp, decline_time_mod_hyp)

head(decline_predict_mod_hyp, 10)

time_abnd <- decline_predict_mod_hyp$`time_abnd_(days)`[1]

EUR <- decline_predict_mod_hyp$`EUR_(bbl)`[1]

p1 <- decline_predict_mod_hyp %>% ggplot(aes(x = `Time_(day)`, y = `q_(bbl/day)`)) + 
   geom_point(color = "red") +
  geom_vline(aes(xintercept = `time_abnd_(days)`), linetype = 2) + 
  annotate(geom = "text", x = time_abnd + 200, y = 1000, label = "time_abnd", angle = 90, 
           color = "blue") +
  geom_hline(aes(yintercept = 5 * 6.289814), linetype = 2) + 
  annotate(geom = "text", x = 8000, y = 200, label = "rate_abnd", color = "blue") +
  theme_bw()

p2 <- decline_predict_mod_hyp %>% ggplot(aes(x = `Time_(day)`, y = `Q_(bbl)`)) + 
  geom_point(color = "red") +
  geom_vline(aes(xintercept = `time_abnd_(days)`), linetype = 2) + 
  annotate(geom = "text", x = time_abnd + 200, y = 3e6, label = "time_abnd", angle = 90, 
           color = "blue") +
  geom_hline(aes(yintercept = `EUR_(bbl)`), linetype = 2) + 
  annotate(geom = "text", x = 8000, y = 1.03 * EUR, label = "EUR", color = "blue") +
  theme_bw()

hyp_plots <- ggarrange(p1, p2, ncol = 1, nrow = 2, align = "v")

hyp_plots

```


#### Example 2:

```{r, fig.width = 6, fig.height= 8, fig.align = "center", warning = TRUE}
library(Rdca)
library(ggplot2)
library(magrittr)

dcl_time_mod_hyp <- decline_time(1:300, unit = "month")

dcl_param_mod_hyp <- decline_param(input_unit = "Field", output_unit = "Field", fluid = "oil", 
                               model = "modified_hyperbolic", qi = 8000, Di = 0.04, b = 0.75, 
                               Dt = 0.01, q_abnd = 10)

set.seed(4321)

prod_data <- decline_predict(dcl_param_mod_hyp, dcl_time_mod_hyp)$`q_(bbl/month)` +
  rnorm(300, mean = 100, sd = 50)   # rate

field_data <- data.frame(time = dcl_time_mod_hyp$t, q = prod_data, Q = cumsum(prod_data))

dcl_fit_param_mod_hyp <- decline_fit_param(input_unit = "Field", output_unit = "Field", fluid = "oil", 
                              model = "modified_hyperbolic", fit_data = "cum", prod_data = field_data$Q, 
                              initial_param = c(10000, 0.1, 1.0, 0.01), lower = NULL,
                              upper = NULL, control = list(maxiter = 100))

tibble::glimpse(dcl_fit_param_mod_hyp)

dcl_fit_mod_hyp <- decline_fit(dcl_fit_param_mod_hyp, dcl_time_mod_hyp)

tibble::glimpse(dcl_fit_mod_hyp)

names(attr(dcl_fit_mod_hyp, which = "nls.out"))

attr(dcl_fit_mod_hyp, which = "nls.out")$par

attr(dcl_fit_mod_hyp, which = "nls.out")$info

attr(dcl_fit_mod_hyp, which = "nls.out")$niter

attr(dcl_fit_mod_hyp, which = "nls.out")$deviance

dcl_predict_mod_hyp <- decline_predict(dcl_fit_mod_hyp, dcl_time_mod_hyp)

p_cum <- field_data %>% ggplot(aes(x = time, y = Q)) +
  geom_point(color = "blue", shape = 21, size = 3) +
  geom_line(aes(x = `Time_(month)`, y = `Q_(bbl)`), data = dcl_predict_mod_hyp, 
            color = "red", size = 1) +
  labs(x = "Time (months)", y = "Cumulative Production (bbl)") + 
  theme_bw()

p_cum

p_rate <- field_data %>% ggplot(aes(x = time, y = q)) +
  geom_point(color = "blue", shape = 21, size = 3) +
  geom_line(aes(x = `Time_(month)`, y = `q_(bbl/month)`), data = dcl_predict_mod_hyp, 
            color = "red", size = 1) +
  labs(x = "Time (months)", y = "Production Rate (bbl/month)") + 
  theme_bw()

p_rate

```

## `References`
