---
title: "Semi-Confirmatory Structural Equation Modeling"
author: "Po-Hsien Huang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Semi-Confirmatory Structural Equation Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r comment = "", message = FALSE, setup, include=FALSE}
options(digits = 3)
options(width = 100)
```
In this example, we will show how to use `lslx` to conduct semi-confirmatory structural equation modeling.
The example uses data `PoliticalDemocracy` in the package `lavaan`.
Hence, `lavaan` must be installed.

## Model Specification
In the following specification, `x1` - `x3` and `y1` - `y8` is assumed to be measurements of 3 latent factors: `ind60`, `dem60`, and `dem65`.
```{r comment = "", message = FALSE}
model_sem <- "fix(1) * x1 + x2 + x3      <=: ind60
              fix(1) * y1 + y2 + y3 + y4 <=: dem60
              fix(1) * y5 + y6 + y7 + y8 <=: dem65
              dem60 <= ind60
              dem65 <= ind60 + dem60"
```
The operator `<=:` means that the RHS latent factors is defined by the LHS observed variables.
In particular, the loadings are freely estimated.
In this model, `ind60` is measured by `x1` - `x3`, `dem60` is mainly measured by `y1` - `y4`, and `dem65` is mainly measured by `y5` - `y8`.
The operator `<=` means that the regression coefficients from the RHS variables to the LHS variables are freely estimated.
In this model, `dem60` is influenced by `ind60` and `dem65` is influenced by `dem60` and `ind60`.
Details of model syntax can be found in the section of Model Syntax via `?lslx`.

## Object Initialization
`lslx` is written as an `R6` class.
Everytime we conduct analysis with `lslx`, an `lslx` object must be initialized.
The following code initializes an `lslx` object named `lslx_sem`.
```{r comment = "", message = FALSE}
library(lslx)
lslx_sem <- lslx$new(model = model_sem,
                    sample_cov = cov(lavaan::PoliticalDemocracy),
                    sample_size = nrow(lavaan::PoliticalDemocracy))
```
Here, `lslx` is the object generator for `lslx` object and `$new()` is the build-in method of `lslx` to generate a new `lslx` object.
The initialization of `lslx` requires users to specify a model for model specification (argument `model`) and a sample moments to be fitted (argument `sample_cov` and `sample_size`).
The sample moment must contain all the observed variables specified in the given model.


## Model Respecification
After an `lslx` object is initialized, model can be respecified by `$free_coefficient()`, `$fix_coefficient()`, and `$penalize_coefficient()` methods.
The following code sets `y1<->y5`, `y2<->y4`, `y2<->y6`, `y3<->y7`, `y4<->y8`, and `y6<->y8` as penalized parameters.
```{r comment = "", message = FALSE}
lslx_sem$penalize_coefficient(name = c("y1<->y5",
                                       "y2<->y4",
                                       "y2<->y6",
                                       "y3<->y7",
                                       "y4<->y8",
                                       "y6<->y8"))
```
To see more methods for respecifying model, please check the section of Set-Related Method via `?lslx`. 


## Model Fitting
After an `lslx` object is initialized, method `$fit_lasso()` can be used to fit the specified model into the given data with LASSO penalty function.
```{r comment = "", message = FALSE}
lslx_sem$fit_lasso()
```
The `$fit_lasso()` requires users to specify the considered penalty levels (argument `lambda_grid`).
In this example, the lambda grid is automatically initialized by default.
Note that MCP with `delta = Inf` is equivalent to the LASSO penalty.
All the fitting result will be stored in the `fitting` field of `lslx_sem`.


## Model Summarizing
Unlike traditional SEM analysis, `lslx` fit the model into data under all the penalty levels considered.
To summarize the fitting result, a selector to determine an optimal penalty level must be specified.
Available selectors can be found in the section of Penalty Level Selection via `?lslx`.
The following code summarize the fitting result under the penalty level selected by adjusted Bayesian information criterion (ABIC).
```{r comment = "", message = FALSE, fig.width = 24, fig.height = 14}
lslx_sem$summarize(selector = "abic")
```
In this example, we can see that the PL estimate under the selected penalty level doesn't contain any zero value, which indicates that all of the covariance of measurements are relevant.
The `$summarize()` method also shows the result of significance tests for the coefficients.
In `lslx`, the default standard errors are calculated based on sandwich formula whenever raw data is available.
In this example, because raw data is not used for `lslx` object initialization, standard error is calculated by using observed Fisher information matrix.
It may not be valid when the model is misspecified and the data are not normal.
Also, it is generally invalid after choosing a penalty level.


## Objects Extraction
In `lslx`, many quantities related to SEM can be extracted by extract-related method.
For example, the coefficient estimate and its asymptotic variance can be obtained by
```{r comment = "", message = FALSE, fig.width = 8, fig.height = 4, dpi=300, out.width=600, out.height=300}
lslx_sem$extract_coefficient(selector = "abic", type = "effective")
```
```{r comment = "", message = FALSE, fig.width = 8, fig.height = 4, dpi=300, out.width=600, out.height=300}
diag(lslx_sem$extract_coefficient_acov(selector = "abic", type = "effective"))
```
Here, the `type` argument is used to specify which types of parameters are used to calculate related quantities. `type = "effective"` indicates that only freely estimated and penalized non-zero parameters are used. By default, `type = "all"`
