---
title: "1 - Running a simple simulation"
data: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1 - Running a simple simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, results="hide", message=FALSE}
library(landsepi)
```

## General presentation of the package

See `?lansdepi` for a complete description of the model, assumptions and available functions. 
See also the [detailed list of parameters](list_of_parameters.pdf) for a detailed description of all model parameters.

## Initialisation of simulation parameters

The function `createSimulParams()` will create a directory to store simulation outputs, 
and return an object of class `LandsepiParams` that will further contain all simulation parameters. 
```{r, results="hide", message="FALSE"}
simul_params <- createSimulParams(outputDir = getwd())
```

In the following, the object `simul_params` is to be updated with all simulation parameters using `set*()` functions. 
To help parameterisation, built-in parameters are available and may be loaded via `load*()` functions. 

## Setting the seed and time parameters

A seed (for random number generator) is randomly generated when `simul_params` is initialised by 
`createSimulParams()`. If a specific seed is required, it can be set using the function `setSeed()`:
```{r}
simul_params@Seed
simul_params <- setSeed(simul_params, seed = 1)
simul_params@Seed
```

The number of cropping seasons to simulate (e.g. 10 years) and the number of timesteps per 
cropping season (e.g. 120 days/year) can be set using `setTime()`:
```{r}
simul_params <- setTime(simul_params, Nyears = 6, nTSpY = 120)
simul_params@TimeParam
```


## Setting pathogen parameters

Pathogen parameters must be stored in a list of aggressiveness components defined on a susceptible host for a 
pathogen genotype not adapted to resistance.  

Buit-in parameterisation for rust pathogens of cereal crops is available using function `loadPathogen()`:
```{r}
basic_patho_param <- loadPathogen(disease = "rust")
basic_patho_param
```

This list may be updated to change a specific parameter. For instance, to change the infection rate to 50%:
```{r}
basic_patho_param <- loadPathogen("rust")
basic_patho_param$infection_rate <- 0.5
basic_patho_param
```

Finally, the list may be generated manually to control all parameters: 
```{r}
basic_patho_param <- list(infection_rate = 0.4
                          , latent_period_exp = 10
                          , latent_period_var = 9
                          , propagule_prod_rate = 3.125
                          , infectious_period_exp = 24
                          , infectious_period_var = 105
                          , survival_prob = 1e-4
                          , repro_sex_prob = 0
                          , sigmoid_kappa = 5.333, sigmoid_sigma = 3, sigmoid_plateau = 1)
```

Then, the list is used to fuel the object `simul_params` via the function `setPathogen()`:
```{r}
simul_params <- setPathogen(simul_params, patho_params = basic_patho_param)
simul_params@Pathogen
```

## Setting inoculum

The function `setInoculum()` updates `simul_params` with the probability for an individual to be infectious 
(i.e. state I) at the beginning of the simulation. Note that the inoculum is present on the first cultivar 
only (usually defined as the susceptible cultivar) and is not adapted to any plant resistance gene. 
```{r}
simul_params <- setInoculum(simul_params, val = 5e-4)
simul_params@PI0
```

## Setting the landscape and pathogen dispersal

The landscape (i.e. boundaries of fields) must be in shapefile format. Five built-in landscapes of about 150 
fields are available using the function `loadLandscape()`:
```{r}
landscape <- loadLandscape(id = 1)
length(landscape)
plot(landscape, main = "Landscape structure")
```

*See also tutorial on how to [parameterise landscape and dispersal](landscape_dispersal.html) 
to use your own landscape and compute your own dispersal matrices.*

Dispersal is given by a vectorised matrix giving the probability of dispersal from any field of the 
landscape to any other field. The size of the matrix must be the square of the number of fields in the landscape. 
It is thus specific to both the pathogen and the landscape. For rusts pathogens, 
a built-in dispersal matrix is available for each landscape using the function `loadDispersalPathogen()`:
```{r}
disp_patho <- loadDispersalPathogen(id = 1)
head(disp_patho)
length(landscape)^2 == length(disp_patho)
```

Then, the object `simul_params` is updated with the landscape and dispersal matrix via the 
functions `setLandscape()` and `setDispersalPathogen()`, respectively:
```{r}
simul_params <- setLandscape(simul_params, land = landscape)
simul_params <- setDispersalPathogen(simul_params, mat = disp_patho)
```

## Setting croptypes, cultivars and resistance genes

Fields of the landscape are cultivated with different croptypes that can rotate through time; 
each croptype is composed of a pure cultivar or a mixture; and each cultivar may carry one or 
several resistance genes.

### Cultivars

Characteristics of each host genotype (i.e. cultivar) are summarised in a dataframe, which 
contains parameters representing the cultivar as if it was cultivated in a pure crop. 
**Note that the name of the cultivar cannot accept spaces.**  
A buit-in parameterisation for classical types of cultivars is available using `loadCultivar()` 
to generate each line of the dataframe. The only difference between the types "growingHost" and 
"nongrowingHost" is the absence of growth in the later. Type "nonCrop" allows the simulation of 
forest, fallows, etc. i.e. everything that is not planted, does not cost anything and does not 
yield anything (with regard to the subject of the study).  
**It is advised to implement a susceptible cultivar at first line of the dataframe to allow 
pathogen initial inoculation.**
```{r}
cultivar1 <- loadCultivar(name = "Susceptible", type = "growingHost")
cultivar2 <- loadCultivar(name = "Resistant1", type = "growingHost")
cultivar3 <- loadCultivar(name = "Resistant2", type = "growingHost")
cultivar4 <- loadCultivar(name = "Resistant3", type = "nongrowingHost")
cultivar5 <- loadCultivar(name = "Forest", type = "nonCrop")
cultivars <- data.frame(rbind(cultivar1, cultivar2, cultivar3, cultivar4, cultivar5)
                        , stringsAsFactors = FALSE)
cultivars
```

Similarly as pathogen parameters, characteristics of the cultivars may be updated as required. 
For example, to change the growth rate of the susceptible cultivar:
```{r}
cultivars[cultivars$cultivarName == "Susceptible", "growth_rate"] <- 0.2
cultivars
```

Finally, the dataframe `cultivars` can also be generated entirely from scratch:
```{r}
cultivars_new <- data.frame(cultivarName = c("Susceptible", "Resistant"),
                            initial_density =   c(0.1, 0.2),
                            max_density =       c(2.0, 3.0),
                            growth_rate =       c(0.1, 0.2),
                            reproduction_rate = c(0.0, 0.0),
                            death_rate =        c(0.0, 0.0),
                            yield_H =           c(2.5, 2.0),
                            yield_L =           c(0.0, 0.0),
                            yield_I =           c(0.0, 0.0),
                            yield_R =           c(0.0, 0.0),
                            production_cost =   c(225, 300),
                            market_value =      c(200, 150),
                            stringsAsFactors = FALSE)
cultivars_new
```

### Resistance genes

Characteristics of each plant resistance gene and each corresponding pathogenicity gene are summarised in a dataframe.  
A built-in parameterisation for classical resistance sources is available using `loadGene()` to generate each line of the 
dataframe. Type "majorGene" is for completely efficient resistance to which pathogen may adapt via a single mutation. Type "APR" 
stands for Adult Plant Resistance, i.e. a major gene which activates after a delay after planting, and type "QTL" is a 
partially efficient resistance source to which the pathogen may adapt gradually. The type "immunity" code for a completely 
efficient resistance gene that cannot be overcome, in such a way that infection is totally impossible; it is helpful 
to parameterise non-host cultivars. 
```{r}
gene1 <- loadGene(name = "MG 1", type = "majorGene")
gene2 <- loadGene(name = "Lr34", type = "APR")
gene3 <- loadGene(name = "gene 3", type = "QTL")
gene4 <- loadGene(name = "nonhost resistance", type = "immunity")
genes <- data.frame(rbind(gene1, gene2, gene3, gene4), stringsAsFactors = FALSE)
genes
```

Similarly as pathogen parameters, characteristics of the genes may be updated as required. For example, to change the 
mutation probability of the pathogen with regard to its adaptation to "MG 1":
```{r}
genes[genes$geneName == "MG 1", "mutation_prob"] <- 1e-3
genes
```

Finally, the dataframe `genes` can also be generated entirely from scratch:
```{r}
genes_new <- data.frame(geneName =               c("MG1", "MG2"),
                        efficiency =             c(1.0  , 0.8  ),
                        time_to_activ_exp =      c(0.0  , 0.0  ),
                        time_to_activ_var =      c(0.0  , 0.0  ),
                        mutation_prob =          c(1E-7 , 1E-4),
                        Nlevels_aggressiveness = c(2    , 2    ),
                        fitness_cost =           c(0.50 , 0.75 ),
                        tradeoff_strength =      c(1.0  , 1.0  ),
                        target_trait =           c("IR" , "LAT"),
                        stringsAsFactors = FALSE)
genes_new
```

Note that the column "efficiency" refers to the percentage of reduction of the targeted aggressiveness 
component on hosts carrying the gene. 
For example, a 80% efficient resistance against infection rate (target_trait = "IR") 
means that the infection rate of a non-adapted pathogen is reduced by 80% on a resistant host compared 
to what it is on a susceptible host. For resistances targetting the latent period, 
the percentage of reduction is applied to the inverse of the latent period in such a way that latent 
period is higher in resistant than in susceptible hosts.


### Allocating genes to cultivars

The object `simul_params` can be updated with `setGenes()` and `setCultivars()`:
```{r}
simul_params <- setGenes(simul_params, dfGenes = genes)
simul_params <- setCultivars(simul_params, dfCultivars = cultivars)
simul_params@Genes
simul_params@Cultivars
```

Then the function `allocateCultivarGenes()` allows the attribution of resistance genes to cultivars:
```{r}
simul_params <- allocateCultivarGenes(simul_params
                                      , cultivarName = "Resistant1"
                                      , listGenesNames = c("MG 1"))
simul_params <- allocateCultivarGenes(simul_params
                                      , cultivarName = "Resistant2"
                                      , listGenesNames = c("Lr34", "gene 3"))
simul_params <- allocateCultivarGenes(simul_params
                                      , cultivarName = "Resistant3"
                                      , listGenesNames = c("nonhost resistance"))
simul_params@Cultivars
```

With this example of parameterisation:  
- "Susceptible" is a susceptible cultivar (initially infected by a wild-type pathogen)  
- "Resistant1" is a mono-resistant cultivar  
- "Resistant2" is a pyramided cultivar  
- "Resistant3" is a nonhost cultivar  
- "Forest" is not a crop.  
Infection is impossible on both "Resistance3" and "Forest", but the former will 
be considered for host growth, yield, production costs and benefits, whereas the latter won't.


### Allocating cultivars to croptypes

Characteristics of each croptype (a croptype is a set of hosts cultivated in a field with specific proportions) 
are summarised in a dataframe.  
A buit-in parameterisation for classical croptypes is available 
using `loadCroptypes()` to generate the whole table filled with zeros: 
```{r}
croptypes <- loadCroptypes(simul_params, names = c("Susceptible crop"
                                                   , "Pure resistant crop"
                                                   , "Mixture"
                                                   , "Other"))
croptypes
```

Then `croptype` is updated by `allocateCroptypeCultivars()` 
to specify the composition of every croptype with regard to cultivars (and proportion for mixtures):
```{r}
croptypes <- allocateCroptypeCultivars(croptypes
                                       , croptypeName = "Susceptible crop"
                                       , cultivarsInCroptype = "Susceptible")
croptypes <- allocateCroptypeCultivars(croptypes
                                       , croptypeName = "Pure resistant crop"
                                       , cultivarsInCroptype = "Resistant1")
croptypes <- allocateCroptypeCultivars(croptypes
                                       , croptypeName = "Mixture"
                                       , cultivarsInCroptype = c("Resistant2","Resistant3")
                                       , prop = c(0.4, 0.6))
croptypes <- allocateCroptypeCultivars(croptypes
                                       , croptypeName = "Other"
                                       , cultivarsInCroptype = "Forest")
croptypes
```

Finally the object `simul_params` is updated using `setCroptypes()`:
```{r}
simul_params <- setCroptypes(simul_params, dfCroptypes = croptypes)
simul_params@Croptypes
```

Alternatively, the dataframe `croptypes` can be generated from scratch:
```{r}
croptypes <- data.frame(croptypeID = c(0, 1, 2, 3)
                        , croptypeName = c("Susceptible crop"
                                           , "Pure resistant crop"
                                           , "Mixture"
                                           , "Other")
                        , Susceptible = c(1,0,0  ,0)
                        , Resistant1  = c(0,1,0  ,0)
                        , Resistant2  = c(0,0,0.5,0)
                        , Resistant3  = c(0,0,0.5,0)
                        , Forest      = c(0,0,0  ,1)
                        , stringsAsFactors = FALSE)
simul_params <- setCroptypes(simul_params, croptypes)
```

### Allocating croptypes to fields of the landscape

The function `allocateLandscapeCroptypes()` manages the allocation of croptypes in time (for rotations 
of different croptypes on the same fields) and space (for mosaic of fields cultivated with different croptypes). 
See `?allocateLandscapeCroptypes` for help in parameters.  
Briefly, a rotation sequence is defined by a list. Each 
element of this list is a vector containing the indices of the croptypes that are cultivated simultaneously in the 
landscape. The "rotation_period" parameter defines the duration before switching from one element of "rotation_sequence" 
to the next. For each element of "rotation_sequence" is associated a vector of proportions of each croptype in the 
landscape (parameter "prop"). The parameter "aggreg" controls the level of spatial aggregation between croptypes. 

For example, to generate a landscape whose surface is composed of 
1/3 of forests, 1/3 of susceptible crop, and 1/3 of fields where a pure resistant crop is alternated with a mixture 
every two years:
```{r}
# croptypeIDs cultivated in each element of the rotation sequence:
rotation_sequence <- list(c(0,1,3), c(0,2,3))
rotation_period <- 2  # number of years before rotation of the landscape
prop <- list(rep(1/3, 3), rep(1/3, 3))  # proportion (in surface) of each croptype
aggreg <- 1    # level of spatial aggregation
simul_params <- allocateLandscapeCroptypes(simul_params
                                           , rotation_period = rotation_period
                                           , rotation_sequence = rotation_sequence
                                           , prop = prop
                                           , aggreg = aggreg
                                           , graphic = FALSE)
# plot(simul_params@Landscape)
```

## Choosing output variables
Several epidemiological, evolutionary and economic outputs can be generated by landsepi and represented in 
text files (`writeTXT=TRUE`), graphics (`graphic=TRUE`) and a video (`videoMP4=TRUE`).  
See `?epid_output` and `?evol_output` for details on the different output variables.  

>Possible epidemiological outputs include:  
- **"audpc"** : Area Under Disease Progress Curve (average proportion of diseased hosts relative 
to the carryng capacity)  
- **"gla_abs"** : Absolute Green Leaf Area (average number of healthy hosts per square meter)  
- **"gla_rel"** : Relative Green Leaf Area  (average proportion of healthy hosts relative to the 
total number of existing hosts)  
- **"eco_cost"** : total crop costs (in weight or volume units per ha)  
- **"eco_product"** : total crop production (in monetary units per ha)  
- **"eco_benefit"** : total crop benefits (in monetary units per ha)  
- **"eco_grossmargin"** : Gross Margin (benefits - costs, in monetary units per ha)  
- **"HLIR_dynamics", "H_dynamics", "L_dynamics", "IR_dynamics", "HLI_dynamics"**, etc.: Epidemic dynamics 
related to the specified sanitary status (H, L, I or R and all their combinations). Graphics only, 
works only if graphic=TRUE.  
- **"all"** : compute all these outputs (default)  
- **""** : none of these outputs will be generated.  

>Possible evolutionary outputs include:  
- **"evol_patho"**: Dynamics of pathogen genotype frequencies  
- **"evol_aggr"**: Evolution of pathogen aggressiveness  
- **"durability"**: Durability of resistance genes  
- **"all"**: compute all these outputs (default)  
- **""**: none of these outputs will be generated.  

A list of outputs can be generated using `loadOutputs()`:
```{r}
outputlist <- loadOutputs(epid_outputs = "all", evol_outputs = "all")
outputlist
```
Then `simul_params` can be updated via `setOutputs()`:
```{r}
simul_params <- setOutputs(simul_params, outputlist)
```

*See also tutorial on how to [run a numerical experimental design](run_exp_design.html) 
to compute your own output variables and to run several simulations within an experimental design*

## Running the simulation

The functions `checkSimulParams()` and `saveDeploymentStrategy()` check simulation parameters and 
save the object `simul_params` (which contains all parameters associated with the deployment strategy) 
into a GPKG file, respectively. 
```{r, eval=FALSE}
checkSimulParams(simul_params)
simul_params <- saveDeploymentStrategy(simul_params)
```

Then, the function `runSimul()` launches the simulation. Use `?runSimul` to get all available options.
```{r, eval=FALSE}
runSimul(simul_params, graphic = TRUE, videoMP4 = FALSE)
```

```{r, include=FALSE}
system(paste("rm -rf ", simul_params@OutputDir))
```


