---
title: "Linear algebra with the `clifford` package"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{linear algebra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This short document shows a nice application of Clifford algebras to
linear algebra.  Suppose we have vectors $a, b, c$ spanning
$x\in\mathcal{R}^3$ and wish to write $x=\alpha a+\beta b+\gamma c$.
The traditional Cramer's rule would be

\[
\alpha=\frac{
	\det\begin{bmatrix}
	x_1&b_1&c_1\\
	x_2&b_2&c_2\\
	x_3&b_3&c_3
	\end{bmatrix}
	}{
	\det\begin{bmatrix}
	a_1&b_1&c_1\\
	a_2&b_2&c_2\\
	a_3&b_3&c_3
	\end{bmatrix}
}
\qquad\beta=\frac{
	\det\begin{bmatrix}
	a_1&x_1&c_1\\
	a_2&x_2&c_2\\
	a_3&x_3&c_3
	\end{bmatrix}
	}{
	\det\begin{bmatrix}
	a_1&b_1&c_1\\
	a_2&b_2&c_2\\
	a_3&b_3&c_3
	\end{bmatrix}
}
\qquad
\gamma=\frac{
	\det\begin{bmatrix}
	a_1&b_1&x_1\\
	a_2&b_2&x_2\\
	a_3&b_3&x_3
	\end{bmatrix}
	}{
	\det\begin{bmatrix}
	a_1&b_1&c_1\\
	a_2&b_2&c_2\\
	a_3&b_3&c_3
	\end{bmatrix}
}.
\]

However, observe that this solution, while accurate, requires one to
take a coordinate basis; and offers little in the way of intuition.

## Using Clifford algebra

Considering $\mathcal{R}^3$ as a vector space and given vectors $a, b,
c$ spanning the space we can express any vector $x\in\mathcal{R}^3$ as

\[
\left(\frac{x\wedge b\wedge c}{a\wedge b\wedge c}\right)a+
\left(\frac{a\wedge x\wedge c}{a\wedge b\wedge c}\right)b+
\left(\frac{a\wedge b\wedge x}{a\wedge b\wedge c}\right)c
\]

which is Cramer's rule expressed directly in vector form (rather than
components).  Package idiom is straightforward:

```{r,results="hide",include=FALSE}
library("clifford",quietly=TRUE)
```

```{r,eval=FALSE}
library("clifford",quietly=TRUE)

```{r,eval=TRUE}
a <- as.1vector(runif(3))
b <- as.1vector(runif(3))
c <- as.1vector(runif(3))

(x <- as.1vector(1:3))
abc <- a %^% b %^% c

xabc <-
a*(x %^% b%^% c)/abc  + 
b*(a %^% x%^% c)/abc  + 
c*(a %^% b%^% x)/abc  
```

Thus we have expressed $x$ as a linear combination of $a,b,c$.  To verify:

```{r}
print(Mod(x-xabc))
```



Conversely, we might know the coefficients [here we will use
$1,2,3$] and try to determine them using package idiom:

```{r}
y <- a*1 + b*2 + c*3
c(
drop((y %^% b %^% c)/abc),
drop((a %^% y %^% c)/abc),
drop((a %^% b %^% y)/abc))
```


# Higher dimensional space

To accomplish this in arbitrary-dimensional space:

```{r}
n <- 13                                                # dimensionality of space
x <- as.1vector(seq_len(n))                            # target vector
L <- replicate(n,as.1vector(rnorm(n)),simplify=FALSE)  # spanning vectors
subst <- function(L,n,x){L[[n]] <- x; return(L)}       # list substitution
coeff <- function(n,L,x){
      drop(Reduce(`%^%`,subst(L,n,x))/Reduce(`%^%`,L))
}
```

Then the coefficients are given by:

```{r}
(alpha <- sapply(seq_len(n),coeff,L,x))
```

and we can reconstitute vector $x$:


```{r}
out <- as.clifford(0)
f <- function(i){alpha[i]*L[[i]]}
for(i in seq_len(n)){
      out <- out + f(i)
}
Mod(out-x)  # zero to numerical precision
```

Or, somewhat slicker:

```{r}
Reduce(`+`,sapply(seq_len(n),f,simplify=FALSE))
```

