---
title: 'Alakazam: Amino acid physicochemical property analysis'
author: "Susanna Marquez"
date: '`r Sys.Date()`'
output:
  pdf_document:
    dev: pdf
    fig_height: 4
    fig_width: 7.5
    highlight: pygments
    toc: yes  
  md_document:
    fig_height: 4
    fig_width: 7.5
    preserve_yaml: no
    toc: yes
  html_document:
    fig_height: 4
    fig_width: 7.5
    highlight: pygments
    theme: readable
    toc: yes
geometry: margin=1in
fontsize: 11pt
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Amino acid property analysis}
  %\usepackage[utf8]{inputenc}
---

The `alakazam` package includes a set of functions to analyze the physicochemical 
properties of Ig and TCR amino acid sequences. Of particular interest is the analysis of 
CDR3 properties, which this vignette will demonstrate. The same process 
can be applied to other regions simply by altering the sequence data column used.

    Wu YC, et al. High-throughput immunoglobulin repertoire analysis distinguishes 
        between human IgM memory and switched memory B-cell populations. 
        Blood 116, 1070-8 (2010).
    Wu YC, et al. The relationship between CD27 negative and positive B cell 
        populations in human peripheral blood. 
        Front Immunol 2, 1-12 (2011).


## Example data

A small example AIRR database, `ExampleDb`, is included in the `alakazam` package.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
# Load required packages
library(alakazam)
library(dplyr)

# Subset example data
data(ExampleDb)
db <- ExampleDb[ExampleDb$sample_id == "+7d", ]
```

For details about the AIRR format, visit the AIRR Community documentation site https://docs.airr-community.org/en/latest/datarep/rearrangements.html#fields

## Calculate the properties of amino acid sequences

Multiple amino acid physicochemical properties can be obtained with the function 
`aminoAcidProperties`. The available properties are:

* `length`: total amino acid count
* `gravy`: grand average of hydrophobicity
* `bulkiness`: average bulkiness
* `polarity`: average polarity
* `aliphatic`: normalized aliphatic index
* `charge`: normalized net charge
* `acidic`: acidic side chain residue content
* `basic`: basic side chain residue content
* `aromatic`: aromatic side chain content

This example demonstrates how to calculate all of the available amino acid 
properties from DNA sequences found in the `junction` column of the previously loaded AIRR file.

Translation of the DNA sequences to amino acid sequences is accomplished by 
specifying the `nt=TRUE` argument. To reduce the junction sequence to the CDR3 
sequence we specify the argument `trim=TRUE` which will strip the first and last
codon (the conserved residues) prior to analysis. The prefix `cdr3` is added
to the output column names using the `label="cdr3"` argument.

```{r, eval=TRUE, warning=FALSE, fig.width=7.5, fig.height=6}
db_props <- aminoAcidProperties(db, seq="junction", nt=TRUE, trim=TRUE, 
                                label="cdr3")

# The full set of properties are calculated by default
dplyr::select(db_props[1:3, ], starts_with("cdr3"))

# Define a ggplot theme for all plots
tmp_theme <- theme_bw() + theme(legend.position="bottom")

# Generate plots for all four of the properties
g1 <- ggplot(db_props, aes(x=c_call, y=cdr3_aa_length)) + tmp_theme +
    ggtitle("CDR3 length") + 
    xlab("Isotype") + ylab("Amino acids") +
    scale_fill_manual(name="Isotype", values=IG_COLORS) +
    geom_boxplot(aes(fill=c_call))
g2 <- ggplot(db_props, aes(x=c_call, y=cdr3_aa_gravy)) + tmp_theme + 
    ggtitle("CDR3 hydrophobicity") + 
    xlab("Isotype") + ylab("GRAVY") +
    scale_fill_manual(name="Isotype", values=IG_COLORS) +
    geom_boxplot(aes(fill=c_call))
g3 <- ggplot(db_props, aes(x=c_call, y=cdr3_aa_basic)) + tmp_theme +
    ggtitle("CDR3 basic residues") + 
    xlab("Isotype") + ylab("Basic residues") +
    scale_y_continuous(labels=scales::percent) +
    scale_fill_manual(name="Isotype", values=IG_COLORS) +
    geom_boxplot(aes(fill=c_call))
g4 <- ggplot(db_props, aes(x=c_call, y=cdr3_aa_acidic)) + tmp_theme +
    ggtitle("CDR3 acidic residues") + 
    xlab("Isotype") + ylab("Acidic residues") +
    scale_y_continuous(labels=scales::percent) +
    scale_fill_manual(name="Isotype", values=IG_COLORS) +
    geom_boxplot(aes(fill=c_call))

# Plot in a 2x2 grid
gridPlot(g1, g2, g3, g4, ncol=2)
```

### Obtaining properties individually

A subset of the properties may be calculated using the `property` argument of
`aminoAcidProperties`. For example, calculations may be restricted to only the 
grand average of hydrophobicity (`gravy`) index and normalized net charge 
(`charge`) by specifying `property=c("gravy", "charge")`. 

```{r, eval=TRUE, warning=FALSE}
db_props <- aminoAcidProperties(db, seq="junction", property=c("gravy", "charge"),
                                nt=TRUE, trim=TRUE, label="cdr3")
dplyr::select(db_props[1:3, ], starts_with("cdr3"))
```

### Using user defined scales

Each property has a default scale setting, but users may specify alternate scales 
if they wish. The following example shows how to import and use the
Kidera et al, 1985 hydrophobicity scale and the Murrary et al, 2006 pK values from 
the `seqinr` package instead of the defaults for calculating the GRAVY index and 
net charge.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
# Load the relevant data objects from the seqinr package
library(seqinr)
data(aaindex)
data(pK)
h <- aaindex[["KIDA850101"]]$I
p <- setNames(pK[["Murray"]], rownames(pK))
# Rename the hydrophobicity vector to use single-letter codes
names(h) <- translateStrings(names(h), ABBREV_AA)
db_props <- aminoAcidProperties(db, seq="junction", property=c("gravy", "charge"), 
                                nt=TRUE, trim=TRUE, label="cdr3", 
                                hydropathy=h, pK=p)
dplyr::select(db_props[1:3, ], starts_with("cdr3"))
```

### Getting vectors of individual properties

The `aminoAcidProperties` function provides a convenient wrapper for calculating
multiple properties at once from a `data.frame`. If a vector of a specific property is
required this may be accomplished using one of the worker functions:

* `gravy`: grand average of hydrophobicity
* `bulk`: average bulkiness
* `polar`: average polarity
* `aliphatic`: aliphatic index
* `charge`: net charge
* `countPatterns`: counts the occurrence of patterns in amino acid sequences

The input to each function must be a vector of amino acid sequences.

```{r, eval=TRUE, warning=FALSE, message=FALSE}
# Translate junction DNA sequences to amino acids and trim first and last codons
cdr3 <- translateDNA(db$junction[1:3], trim=TRUE)

# Grand average of hydrophobicity
gravy(cdr3)

# Average bulkiness
bulk(cdr3)

# Average polarity
polar(cdr3)

# Normalized aliphatic index
aliphatic(cdr3)
# Unnormalized aliphatic index
aliphatic(cdr3, normalize=FALSE)

# Normalized net charge
charge(cdr3)
# Unnormalized net charge
charge(cdr3, normalize=FALSE)

# Count of acidic amino acids
# Takes a named list of regular expressions
countPatterns(cdr3, c(ACIDIC="[DE]"), label="cdr3")
```

## Default scales

The following references were used for the default physicochemical scales:

* Aliphatic index:  
  Ikai AJ. Thermostability and aliphatic index of globular proteins. 
        J Biochem 88, 1895-1898 (1980).
* Bulkiness scale:  
  Zimmerman JM, Eliezer N, Simha R. The characterization of amino acid sequences 
    in proteins by statistical methods. 
    J Theor Biol 21, 170-201 (1968).
* Hydrophobicity scale:  
  Kyte J, Doolittle RF. A simple method for displaying the hydropathic character 
    of a protein.
    J Mol Biol 157, 105-32 (1982).
* pK values:  
  \url{http://emboss.sourceforge.net/apps/cvs/emboss/apps/iep.html}
* Polarity scale:  
  Grantham R. Amino acid difference formula to help explain protein evolution. 
    Science 185, 862-864 (1974).
