---
title: "Set working directory OS-dependent with `setwdOS`"
author: "Florian Detsch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Set working directory OS-dependent with `setwdOS`}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r pkg, echo=FALSE, message=FALSE}
library(Orcs)
```

Due to work-related,  system exclusive software, I got used to switching from Linux to Windows, and vice versa, quite often. Since rebooting the system under Linux every time I wanted to do a little coding in between seemed inconvenient, I usually included a simple `switch` command in the beginning of my code to identify the current OS (via `Sys.info()[["sysname"]]` and set the working directory accordingly. These few lines of code are now available as a function, which might come in handy for folks switching between different OS as often as I do.

The usage is almost self-explanatory. 

```{r setwdOS, eval=FALSE}
## os partition without file path extension
setwdOS()

## with file path extension
setwdOS(path_ext = "kilimanjaro/nubiscope")
```
