---
title: "Merging Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merging Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BinaryDosage)
```

Quite often subjects have their genotypes imputed in batches. The files returned by these imputation can be converted into binary dosage files. These binary files can be merged into a single file if they have the same SNPs and different subjects using the bdmerge routine.

## bdmerge

The bdmerge routine takes the following parameters

- mergefiles - A character vector of the binary dosage file, family file, and map file names
- format - Integer value indicating which format of the binary dosage file should be used for the merged files
- subformat - Integer value indicating which subformat should be used for the merged files
- bdfiles - A character vector of the binary dosage files to merge
- famfiles - Character vector of the family files associated with the binary dosage files to merge
- mapfiles - Character vector of the map files associated with the binary dosage files to merge
- onegroup - Logical value indicating if the binary dosage saves SNP summary information about each merged file
- bdoptions - Character vector indicating on which SNP information should be evaluated for the merged files. This cannot be used if onegroup is set to FALSE
- snpjoin - Character value indicating if an inner or outer join is done for the SNPs

The following code merges *vcf1a.bdose* and *vcf1b.bdose* into one binary dosage file. It then displays the number of subjects in each file.

``` {r, eval = T, echo = T, message = F, warning = F, tidy = T}
bd1afile <- system.file("extdata", "vcf1a.bdose", package = "BinaryDosage")
bd1bfile <- system.file("extdata", "vcf1b.bdose", package = "BinaryDosage")
bd1file <- tempfile()

bdmerge(mergefiles = bd1file, bdfiles = c(bd1afile, bd1bfile))

bd1ainfo <- getbdinfo(bd1afile)
bd1binfo <- getbdinfo(bd1bfile)
bd1info <- getbdinfo(bd1file)

nrow(bd1ainfo$samples)
nrow(bd1binfo$samples)
nrow(bd1info$samples)
```
